---
title: "eas509_project1"
date: "2024-04-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Install required packages and libraries

```{r}
library(randomForest)
library(factoextra)
library(e1071)
library(readxl)
library(caret)
library(rpart)
library(class)
library(pROC)
```


```{r}
# Load the data, remove the first row and add new column names
mine_data <- read_excel("Mine_Dataset.xls", sheet=2, col_names = FALSE)
first_row <- read_excel("Mine_Dataset.xls", n_max = 1, col_names = FALSE)
mine_data <- mine_data[-1, ]
head(mine_data)
```

Removing the first row and add new column names
Removing the first row and adding new column names

```{r}
mine_data <- mine_data[-1, ]
new_col_names=c("voltage", "sensor_height", "soil_type", "mine_type")
colnames(mine_data) <- new_col_names
head(mine_data)
```

Below are the details regarding the naming

Voltage (V) :Output voltage value of FLC sensor due to magnetic distortion High (H): The height of the sensor from the ground. Soil Type (S):6 different soil types depending on the moisture condition. Mine Type (M): Mine types commonly encountered on land. 5 different mine classes.

```{r}
dim(mine_data)
```

```{r}
# Convert character types to numeric values
mine_data$voltage <- as.numeric(mine_data$voltage)
mine_data$sensor_height <- as.numeric(mine_data$sensor_height)
mine_data$soil_type <- as.numeric(mine_data$soil_type)
mine_data$mine_type <- as.numeric(mine_data$mine_type)
str(mine_data)
```

## EDA
```{r}
#check for null values
sum(is.na(mine_data))

#summary statistics
summary(mine_data)
```

```{r}
#To view the distribution of each numeric variable
hist(mine_data$voltage, main = "Histogram of Voltage", xlab = "Voltage")
hist(mine_data$sensor_height, main = "Histogram of Sensor Height", 
     xlab = "Sensor Height")

# Frequency table and distribution of the target variable mine_type
table(mine_data$mine_type)
barplot(table(mine_data$mine_type), main = "Distribution of Mine Types", 
        xlab = "Mine Type", ylab = "Frequency")
```

#### Outlier Detection

```{r}
boxplot(mine_data$voltage, main = "Boxplot of Voltage", ylab = "Voltage")
#Presence of outliers

boxplot(mine_data$sensor_height, main = "Boxplot of Sensor Height", 
        ylab = "Sensor Height")
# This boxplot of sensor height does not show any individual points above 
# the upper whisker or below the lower whisker, which means there are no apparent 
# outliers
```


#### Checking relationship between variables

```{r}
# Set the size of the plotting area
par(mfrow = c(1, 1), mar = c(5, 1, 2, 1))
plot(mine_data$voltage, mine_data$sensor_height, 
     main = "Voltage vs. Sensor Height", xlab = "Voltage", 
     ylab = "Sensor Height", pch = 19)

# The scatter plot examining the relationship between voltage and sensor height 
# reveals that there is no evident correlation between these two variables.
# Sensor height values are recorded at discrete intervals. While there is 
# considerable variation in voltage measurements at most sensor heights, there 
# does not seem to be a consistent pattern linking voltage to sensor height.
```


#### Checking if there exists relationship between the target variables and numerical variables

```{r}
# Boxplot for voltage by mine type
boxplot(voltage ~ mine_type, data = mine_data, main = "Voltage by Mine Type", 
        xlab = "Mine Type", ylab = "Voltage")
# It seems that as mine type increases, there may be a tendency for both the 
# median voltage and the variability of voltage to increase. 
# The outliers in each category suggest some voltage readings that are unusually 
# high.

# Boxplot for sensor height by mine type
boxplot(sensor_height ~ mine_type, data = mine_data, 
        main = "Sensor Height by Mine Type", xlab = "Mine Type", 
        ylab = "Sensor Height")

# Unlike voltage, sensor height does not show a clear trend as mine type 
# increases.
# All mine types from 1 to 4 have a similar range of sensor heights. 
```

#### Checking if there exists relationship between the target variables and categorical variables

```{r}
# Cross-tabulation with soil type
table(mine_data$soil_type, mine_data$mine_type)
# The cross-tabulation of soil type and mine type reveals a homogenous 
# distribution, indicating no apparent pattern of specific soil types for any 
# given mine type. Each mine type has a similar count across various soil types

#checking if any correlation exist
cor(mine_data[, c("voltage", "sensor_height", "soil_type","mine_type")])

# These correlations suggest that there are no strong linear relationships 
# between the variables in this dataset. 
# The strongest relationship observed is a moderate negative correlation between 
# voltage and sensor height. 
```


```{r}
pairs(mine_data)
```

```{r}
cor(mine_data)
```

```{r}
plot(density(mine_data$voltage))
plot(density(mine_data$sensor_height))
plot(density(mine_data$soil_type))
plot(density(mine_data$mine_type))
```

## MODELS

### 1) PCA

```{r}
pca_result <- prcomp(mine_data[, c("voltage", "sensor_height", "soil_type", 
                                   "mine_type")], center = TRUE, scale. = TRUE)
summary(pca_result)
```

```{r}
plot(pca_result$sdev^2 / sum(pca_result$sdev^2), xlab = "Principal Component",
     ylab = "Proportion of Variance Explained", ylim = c(0, 0.4), type = 'b', 
     main = "Proportion of Variance Explained")
```

Display the cumulative variance explained

```{r}
cumulative_variance_explained <- 
  cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100
plot(cumulative_variance_explained,type = 'b',
     main= "Cumulative Proportion of Variance Explained")

cumulative_variance_df <- data.frame(
  Principal_Component = 1:length(cumulative_variance_explained),
  Cumulative_Variance_Explained = cumulative_variance_explained
)

cumulative_variance_df
```

### 2) K-Means clustering

```{r}
scaled_data <- scale(mine_data[, -ncol(mine_data)])  
# Assuming the last column is 'mine_type', a categorical variable
fviz_nbclust(scaled_data, kmeans, method = "wss")
```

In the plot, the "elbow" point, where there is a noticeable shift in the slope of the curve, is identified at K=5. This point is used as a guideline for determining the most suitable number of clusters. Selecting 5 clusters helps achieve a balance between creating tightly grouped clusters and avoiding the risk of overfitting the model to the data. Therefore, based on this analysis, 5 clusters are deemed optimal for this dataset.
Viewing the summary of number of clusters determined to be optimal

```{r}
k <- 5  
# Perform k-mean clustering with voltage, sensor_height and soil_type
km_result <- kmeans(scaled_data, centers = k, nstart = 25)
print(km_result)
```


```{r}
# Split the data into train and test sets
set.seed(4)
train_index <- createDataPartition(mine_data$mine_type, p = 0.8, list = FALSE)
train_data <- mine_data[train_index, ]
test_data <- mine_data[-train_index, ]

# Convert mine_type to a factor variable
train_data$mine_type <- as.factor(train_data$mine_type)
test_data$mine_type <- as.factor(test_data$mine_type)
```

### 3) Random forest model

```{r}
rf_model <- randomForest(mine_type ~ ., data = train_data)
rf_pred <- predict(rf_model, test_data)
rf_accuracy <- mean(rf_pred == test_data$mine_type)
print(paste("RandomForest Accuracy:", rf_accuracy))

conf_matrix <- confusionMatrix(data = rf_pred, reference = test_data$mine_type)
accuracy <- conf_matrix$overall['Accuracy']
precision <- conf_matrix$byClass[, "Precision"]
recall <- conf_matrix$byClass[, "Recall"]
F1_score <- conf_matrix$byClass[, "F1"]

# Calculate the average precision, recall, and F1-score
avg_precision <- mean(precision, na.rm = TRUE)
avg_recall <- mean(recall, na.rm = TRUE)
avg_F1_score <- mean(F1_score, na.rm = TRUE)

# Print the average precision, recall, and F1-score
print(paste("Average Precision:", avg_precision))
print(paste("Average Recall:", avg_recall))
print(paste("Average F1-score:", avg_F1_score))
```

### 4) Naive bayes model

```{r}
nb_model <- naiveBayes(mine_type ~ ., data = train_data)
nb_pred <- predict(nb_model, test_data, type = "class")
nb_accuracy <- mean(nb_pred == test_data$mine_type)
print(paste("Naive Bayes Accuracy:", nb_accuracy))

conf_matrix <- confusionMatrix(data = nb_pred, reference = test_data$mine_type)
accuracy <- conf_matrix$overall['Accuracy']
precision <- conf_matrix$byClass[, "Precision"]
recall <- conf_matrix$byClass[, "Recall"]
F1_score <- conf_matrix$byClass[, "F1"]

# Calculate the average precision, recall, and F1-score
avg_precision <- mean(precision, na.rm = TRUE)
avg_recall <- mean(recall, na.rm = TRUE)
avg_F1_score <- mean(F1_score, na.rm = TRUE)

# Print the average precision, recall, and F1-score
print(paste("Average Precision:", avg_precision))
print(paste("Average Recall:", avg_recall))
print(paste("Average F1-score:", avg_F1_score))
```

### 5) Decision tree classification

```{r}
tree_model <- rpart(mine_type ~ ., data = train_data, method = "class")
tree_pred <- predict(tree_model, test_data, type = "class")
tree_accuracy <- mean(tree_pred == test_data$mine_type)

conf_matrix <- confusionMatrix(data = tree_pred, reference = test_data$mine_type)
accuracy <- conf_matrix$overall['Accuracy']
precision <- conf_matrix$byClass[, "Precision"]
recall <- conf_matrix$byClass[, "Recall"]
F1_score <- conf_matrix$byClass[, "F1"]

# Calculate the average precision, recall, and F1-score
avg_precision <- mean(precision, na.rm = TRUE)
avg_recall <- mean(recall, na.rm = TRUE)
avg_F1_score <- mean(F1_score, na.rm = TRUE)

# Print the average precision, recall, and F1-score
print(paste("Decision Tree Accuracy:", tree_accuracy))
print(paste("Average Precision:", avg_precision))
print(paste("Average Recall:", avg_recall))
print(paste("Average F1-score:", avg_F1_score))
```

### 6) k-Nearest Neighbors

```{r}
# Define the number of neighbors (k)
k <- 5 
knn_model <- knn(train = train_data[, -ncol(train_data)],  
                 test = test_data[, -ncol(test_data)],    
                 cl = train_data$mine_type,              
                 k = k)                                  

# Measure accuracy
accuracy <- mean(knn_model == test_data$mine_type)
print(paste("kNN Accuracy:", accuracy))

# Calculate confusion matrix
conf_matrix <- confusionMatrix(knn_model, test_data$mine_type)
accuracy <- conf_matrix$overall['Accuracy']
precision <- conf_matrix$byClass[, "Precision"]
recall <- conf_matrix$byClass[, "Recall"]
F1_score <- conf_matrix$byClass[, "F1"]

# Calculate the average precision, recall, and F1-score
avg_precision <- mean(precision, na.rm = TRUE)
avg_recall <- mean(recall, na.rm = TRUE)
avg_F1_score <- mean(F1_score, na.rm = TRUE)

# Print the average precision, recall, and F1-score
print(paste("Average Precision:", avg_precision))
print(paste("Average Recall:", avg_recall))
print(paste("Average F1-score:", avg_F1_score))
```

### 7) Support vector machines

```{r}
# Train the SVM model
svm_model <- svm(mine_type ~ ., data = train_data, kernel = "radial", probability = TRUE)

# Make predictions on the test data
svm_pred <- predict(svm_model, test_data, probability = TRUE)
conf_matrix <- confusionMatrix(data = svm_pred, reference = test_data$mine_type)

# Calculate accuracy
svm_accuracy <- mean(svm_pred == test_data$mine_type)
print(paste("SVM Accuracy:", svm_accuracy))

accuracy <- conf_matrix$overall['Accuracy']
precision <- conf_matrix$byClass[, "Precision"]
recall <- conf_matrix$byClass[, "Recall"]
F1_score <- conf_matrix$byClass[, "F1"]

# Calculate the average precision, recall, and F1-score
avg_precision <- mean(precision, na.rm = TRUE)
avg_recall <- mean(recall, na.rm = TRUE)
avg_F1_score <- mean(F1_score, na.rm = TRUE)

# Print the average precision, recall, and F1-score
print(paste("Average Precision:", avg_precision))
print(paste("Average Recall:", avg_recall))
print(paste("Average F1-score:", avg_F1_score))

svm_pred_prob <- attr(svm_pred, "probabilities")[, "3"]
binary_response <- ifelse(test_data$mine_type == "3", 1, 0)
roc_results <- roc(binary_response, svm_pred_prob)
```





